---
title: "Chapter3notes"
author: "Nicole E Soltis"
date: "February 19, 2016"
output: 
  html_document: 
    keep_md: yes
---
##Example 3.1
```{r}
PrPV <- 0.95
PrPM <- 0.01
PrV <- 0.001
PrP <- PrPV*PrV + PrPM*(1-PrV)
( PrVP <- PrPV*PrV / PrP )
```
## 3.2
```{r}
p_grid <- seq( from=0 , to=1 , length.out=1000 )
prior <- rep( 1 , 1000 )
likelihood <- dbinom( 6 , size=9 , prob=p_grid )
posterior <- likelihood * prior
posterior <- posterior / sum(posterior)
```
## 3.3 random sampling with replacement from the vector p_grid
```{r}
samples <- sample( p_grid , prob=posterior , size=1e4 , replace=TRUE )
```
if not including prob=posterior, will sample randomly

## 3.4
```{r}
plot(samples)
```

## 3.5
```{r}
library(rethinking)
dens(samples)
```

## 3.6
add up posterior probability where p < 0.5 
```{r}
sum( posterior[ p_grid < 0.5 ] )
```

## 3.7
```{r}
sum( samples < 0.5 ) / 1e4
```

## 3.8
```{r}
sum( samples > 0.5 & samples < 0.75 ) / 1e4
```

## 3.9
```{r}
quantile( samples , 0.8 )
```
80% of the data is below this cutoff

## 3.10
```{r}
quantile( samples , c( 0.1 , 0.9 ) )
```

## 3.11
```{r}
p_grid <- seq( from=0 , to=1 , length.out=1000 )
prior <- rep(1,1000)
likelihood <- dbinom( 3 , size=3 , prob=p_grid )
posterior <- likelihood * prior
posterior <- posterior / sum(posterior)
samples <- sample( p_grid , size=1e4 , replace=TRUE , prob=posterior )
```

## 3.12
```{r}
PI( samples , prob=0.5 )
```
percentile interval

## 3.13
```{r}
HPDI( samples , prob=0.5 )
```
highest posterior density interval

narrowest interval

if data is normal or perfectly symmetrical, HPDI is equal to PI

more different if data is skewed

## 3.14
```{r}
p_grid[ which.max(posterior) ]
```

## 3.15
```{r}
chainmode( samples , adj=0.01 )
```

## 3.16
```{r}
mean( samples )
median( samples )
```

## 3.17
```{r}
sum( posterior*abs( 0.5 - p_grid ) )
```

##3.18
```{r}
#d is set to one value of p_grid vector at a time
#and for each value of d, calculates a new vector of difference between d and ALL of p_grid
#figuring out loss for EVERY POSSIBLE decision, d
loss <- sapply( p_grid , function(d) sum( posterior*abs( d - p_grid ) ) )
plot(p_grid,loss)
which.min(loss)
p_grid[which.min(loss)]
```

sapply: take each item in (vector or list), apply function to it

similar to pulling a function down a column in excel
```{r}
square_a_number <- function(x) x*x
sapply(p_grid, square_a_number)

#quicker version
sapply(p_grid, function(x) x*x)
```

##3.19
```{r}
p_grid[ which.min(loss) ]
```

You can also embed plots, for example:

```{r, echo=FALSE}
plot(cars)
```
