# Statistical Rethinking Chapter 4 problems

__Name:__Nicole Soltis


# For 03/17/16

## 4E1
The first line (yi) is the likelihood

## 4E2
The posterior distribution has 2 parameters (mu and sigma).

## 4E3
yi ~ Normal(mu; sigma)
mu ~ Normal(0; 10)
sigma ~ Uniform(0; 10)



\begin{equation}
Pr(mu, sigma|y) = (producti Normal(yi|mu, sigma)* Normal(mu|0,10) * Uniform(sigma|0,10))/ (sum sum numerator) * d mu * d sigma
\end{equation}

## 4M1

Simulate observed heights from the prior of above
```{r}
library(rethinking)
sample_mu <- rnorm( 1e4 , 0 , 10 ) #normal, centered at 0, sd is 10
sample_sigma <- runif( 1e4 , 0 , 10 )
prior <- rnorm( 1e4 , sample_mu , sample_sigma )
dens( prior )
```
## 4M2
now as map formula
```{r}
#make alist of 3 equations to define the model
flist <- alist(
height ~ dnorm( mu , sigma ) ,
mu ~ dnorm( 0 , 10 ) , #mean is 0, sd is 10
sigma ~ dunif( 0 , 10 )
)

```
# For 03/24/16

## 4E4
linear model is: mui = alpha + beta*xi

## 4E5
parameters in posterior: alpha, beta (which model mu with certainty), sigma. So really, 3 parameters.

## 4M3
flist <- alist(
y ~ dnorm( mu , sigma ),
mu <- a + b*x,
a ~ dnorm( 0 , 50 ),
b ~ dunif( 0 , 10 ),
sigma ~ dunif( 0 , 50 )
)

Mathematical model definition:
using R markdown and LaTeX
~ in R markdown makes a space
& aligns lines on the symbol
$$
\begin{align}
y_i &\tilde ~~ Normal(\mu _i, \sigma)\\
\mu _i &= \alpha +  \beta * x_i \\
\alpha &\tilde Normal(0,50)\\
\beta &\tilde Normal(0,10)\\
\sigma &\tilde Uniform(0,50)\\
\end{align}
$$

## 4M4
```{r}
hti ~ dnorm( mu , sigma ),
mu <- a + b*yeari,
a ~ dnorm( 150 , 50 ),
#height in cm
b ~ dunif( 1 , 10 ), #kids grow 1-10 cm/ year?
#slope in height, year
sigma ~ dunif( 0 , 50 )
```
## 4M5
```{r}
#decrease prior for height to match data
#and smaller sd because more informative prior
#data already consistent with height increase annually
ht ~ dnorm( mu , sigma ),
mu <- a + b*year,
a ~ dnorm( 120 , 20 ),
#height in cm
b ~ dunif( 1 , 10 ),
#slope in height, year
sigma ~ dunif( 0 , 50 )
```
## 4M6
```{r}
#variance (sd^2) in heights among cohort never over 64cm
#so I'll use a smaller sigma from 0 to 8
ht ~ dnorm( mu , sigma ),
mu <- a + b*year,
a ~ dnorm( 120 , 20 ),
#height in cm
b ~ dunif( 1 , 10 ),
#slope in height, year
sigma ~ dunif( 0 , 8 )
```
# For 03/31/16

## 4H1
#predicted height and 89% HDPI / PI for each individual

```{r}
library(rethinking)
data(Howell1)
d <- Howell1
model <- map(
  alist(
    height ~ dnorm( mu , sigma ) ,
    mu <- a + b*weight ,
    a ~ dnorm( 178 , 100 ) ,
    b ~ dnorm( 0 , 10 ) ,
    sigma ~ dunif( 0 , 50 )
) , data=d )

#Gina's option
post <- extract.samples(model) #all combinations of a, b when MAP fit the model
unknown_weights=c(46.95,43.72,64.78,32.59,54.63)
mu.link <- function(weight) post$a + post$b*weight

#Julin's option
#standardize weights
#use link
#predict mean height for a given weight
#sim heights for a given weight
#calculate PIs
#print as data frame
#kable from knitr library
```

## 4H2
```{r}
library(rethinking)
data(Howell1)
d <- Howell1
d2 <- d[ d$age < 18 , ]

#a. fit a linear regression
mymap <- map(
  alist(
    height
  )
)
#b. superimpose MAP regression line and HDPI
#c. better model?
```
## 4H3
```{r}
#similar to previous formula but log(weight)
```
