# Statistical Rethinking Chapter 4 problems

__Name:__Nicole Soltis


# For 03/17/16

## 4E1
The first line (yi) is the likelihood

## 4E2
The posterior distribution has 2 parameters (mu and sigma).

## 4E3
yi ~ Normal(mu; sigma)
mu ~ Normal(0; 10)
sigma ~ Uniform(0; 10)



\begin{equation}
Pr(mu, sigma|y) = (producti Normal(yi|mu, sigma)* Normal(mu|0,10) * Uniform(sigma|0,10))/ (sum sum numerator) * d mu * d sigma
\end{equation}

## 4M1

Simulate observed heights from the prior of above
```{r}
library(rethinking)
sample_mu <- rnorm( 1e4 , 0 , 10 )
sample_sigma <- runif( 1e4 , 0 , 10 )
prior_y <- rnorm( 1e4 , sample_mu , sample_sigma )
dens( prior_y )
```
## 4M2
now as map formula
```{r}
#make alist of 3 equations to define the model
flist <- alist(
y ~ dnorm( mu , sigma ) ,
mu ~ dnorm( 0 , 10 ) ,
sigma ~ dunif( 0 , 10 )
)

```
# For 03/24/16

## 4E4
linear model is: mui = alpha + beta*xi

## 4E5
parameters in posterior: alpha, beta (which model mu with certainty), sigma. So really, 3 parameters.

## 4M3
flist <- alist(
y ~ dnorm( mu , sigma ),
mu <- a + b*x,
a ~ dnorm( 0 , 50 ),
b ~ dunif( 0 , 10 ),
sigma ~ dunif( 0 , 50 )
)

Mathematical model definition:
\begin{equation}

\end{equation}
## 4M4

## 4M5

## 4M6

# For 03/31/16

## 4H1
#predicted height and 89% HDPI / PI for each individual

#1: 46.95
#2: 43.72
#3: 64.78
#4: 32.59
#5: 54.63

mN <- map(
alist(
height ~ dnorm( mu , sigma ) ,
mu <- a + b*weight ,
a ~ dnorm( 178 , 100 ) ,
b ~ dnorm( 0 , 10 ) ,
sigma ~ dunif( 0 , 50 )
) , data=dN )


## 4H2

## 4H3
