---
title: "Chapter-07-notes"
author: "Nicole E Soltis"
date: "May 23, 2016"
output: 
  html_document: 
    keep_md: yes
---
For a complex data set, include only certain factors in model
Others- can try adding individual terms to model

Selecting priors:
- alpha (intercept): set prior mean to mean of data, set prior sd to sd of data
- beta (slopes): conservatively set prior mean to 0, means you have no a priori expectation of effect direction. Can eyeball sd from plot of data, choose a relatively broad sd. 
- regularizing priors: if tons of possible predictors and assume most will be 0, keep sds of each beta small (force most to 0). e.g. GWAS: 100k SNPs in one model, want mean of all to be 0, sd to be small --> most have no effect.
- if standardizing variables, alpha should DEFINITELY be mean of the data. And having sd of 1 is reasonable if everything is standardized. 
- range in y / range in x ~= maximum slope estimate
- even if you use really broad priors, evaluating the model by the methods taught here is very useful!

```{r}
#7.1
#data on terrain ruggedness in Africa vs. other continents
#regression: rugged (~ bad geography) against log GDP per capita
library(rethinking)
data(rugged)
d <- rugged
# make log version of outcome
d$log_gdp <- log( d$rgdppc_2000 )
# extract countries with GDP data
dd <- d[ complete.cases(d$rgdppc_2000) , ]
# split countries into Africa and not-Africa
d.A1 <- dd[ dd$cont_africa==1 , ] # Africa
d.A0 <- dd[ dd$cont_africa==0 , ] # not Africa

#7.2 fit the regression models
# African nations
m7.1 <- map(
  alist(
   log_gdp ~ dnorm( mu , sigma ) ,
   mu <- a + bR*rugged ,
   a ~ dnorm( 8 , 100 ) ,
   bR ~ dnorm( 0 , 1 ) ,
   sigma ~ dunif( 0 , 10 )
  ) ,
data=d.A1 )
# non-African nations
m7.2 <- map(
alist(
log_gdp ~ dnorm( mu , sigma ) ,
mu <- a + bR*rugged ,
a ~ dnorm( 8 , 100 ) ,
bR ~ dnorm( 0 , 1 ) ,
sigma ~ dunif( 0 , 10 )
) ,
data=d.A0 )

#plot the posterior predictions
rug.seq <- seq(from=-1,to=6,length.out=30)
d.predict.7.2 <- list(
  log_gdp = rep(0,30), # empty outcome
  rugged = rug.seq
)

#non-African nations
pred.7.2 <- link( m7.2 , data=d.predict.7.2 )
mu.7.2 <- apply( pred.7.2 , 2 , mean )
mu.PI.7.2 <- apply( pred.7.2 , 2, function(x) PI(x, prob=0.97) )
# plot it all
plot( log_gdp ~ rugged , d.A0 , col=rangi2 )
#dashed regression line
lines( rug.seq , mu.7.2 , lty=2 )
shade(mu.PI.7.2, rug.seq)

#African nations
pred.7.1 <- link( m7.1 , data=d.predict.7.2 )
mu.7.1 <- apply( pred.7.1 , 2 , mean )
mu.PI.7.1 <- apply( pred.7.1 , 2, function(x) PI(x, prob=0.97) )
plot( log_gdp ~ rugged , d.A1 , col=rangi2 )
lines( rug.seq , mu.7.1 , lty=2 )
shade(mu.PI.7.1, rug.seq)

#7.3
#how much does singling out nations in Africa change predictions?
#linear regression of log_gdp on ruggedness for entire data set
m7.3 <- map(
alist(
log_gdp ~ dnorm( mu , sigma ) ,
mu <- a + bR*rugged ,
a ~ dnorm( 8 , 100 ) ,
bR ~ dnorm( 0 , 1 ) ,
sigma ~ dunif( 0 , 10 )
) ,
data=dd )

#7.4
#same model but including a dummy variable (slope bA) for nation
m7.4 <- map(
alist(
log_gdp ~ dnorm( mu , sigma ) ,
mu <- a + bR*rugged + bA*cont_africa ,
a ~ dnorm( 8 , 100 ) ,
bR ~ dnorm( 0 , 1 ) ,
bA ~ dnorm( 0 , 1 ) ,
sigma ~ dunif( 0 , 10 )
) ,
data=dd )

#7.5 
#compare these models with WAIC
compare( m7.3 , m7.4 )
#m7.4 gets all the weight- can ignore m7.3

#7.6
#plot the posterior predictions for m7.4
#sample from the posterior and compute the predicted means and intervals
rugged.seq <- seq(from=-1,to=8,by=0.25)
# compute mu over samples, fixing cont_africa=0
mu.NotAfrica <- link( m7.4 , data=data.frame(cont_africa=0,rugged=rugged.seq) )
# compute mu over samples, fixing cont_africa=1
mu.Africa <- link( m7.4 , data=data.frame(cont_africa=1,rugged=rugged.seq) )
# summarize to means and intervals
mu.NotAfrica.mean <- apply( mu.NotAfrica , 2 , mean )
mu.NotAfrica.PI <- apply( mu.NotAfrica , 2 , PI , prob=0.97 )
mu.Africa.mean <- apply( mu.Africa , 2 , mean )
mu.Africa.PI <- apply( mu.Africa , 2 , PI , prob=0.97 )
#all this has done is allow a unique mean prediction (regression mean) depending on the dummy variable cont_africa. It does not allow unique slopes.

#7.7
#fit a new model including an interaction between ruggedness and continent
m7.5 <- map(
alist(
log_gdp ~ dnorm( mu , sigma ) ,
mu <- a + gamma*rugged + bA*cont_africa ,
#model the slope of ruggedness AS A LINEAR MODEL
#in which slope is conditional upon continent
gamma <- bR + bAR*cont_africa ,
#very flat prior for the intercept
#intercept can be regularized by the priors
a ~ dnorm( 8 , 100 ) ,
#weakly regularizing priors for the coefficients
bA ~ dnorm( 0 , 1 ) ,
bR ~ dnorm( 0 , 1 ) ,
bAR ~ dnorm( 0 , 1 ) ,
sigma ~ dunif( 0 , 10 )
) ,
data=dd )
#gamma gets evaluated first, then plugged into mu, then likelihood is computed

#7.8
#compare model with linear interaction to the previous two models
compare( m7.3 , m7.4 , m7.5 )
#STRONG support for including the interaction fx (model has 96% of weight)
#but: some weight to 7.4: suggests posterior means of slopes are slightly overfit
#SE of difference in top 2 models ~ same

#7.9
#can multiply out interaction effects so that there is only one model
#equivalent to model 7.5 above
#and this is more similar to lm() notation
m7.5b <- map(
alist(
log_gdp ~ dnorm( mu , sigma ) ,
mu <- a + bR*rugged + bAR*rugged*cont_africa + bA*cont_africa,
a ~ dnorm( 8 , 100 ) ,
bA ~ dnorm( 0 , 1 ) ,
bR ~ dnorm( 0 , 1 ) ,
bAR ~ dnorm( 0 , 1 ) ,
sigma ~ dunif( 0 , 10 )
) ,
data=dd )

#7.10
#calculate posterior mean line and interval for Africa & not-Africa plots
rugged.seq <- seq(from=-1,to=8,by=0.25) 
mu.Africa <- link( m7.5 , data=data.frame(cont_africa=1,rugged=rugged.seq) )
mu.Africa.mean <- apply( mu.Africa , 2 , mean )
mu.Africa.PI <- apply( mu.Africa , 2 , PI , prob=0.97 )
mu.NotAfrica <- link( m7.5 , data=data.frame(cont_africa=0,rugged=rugged.seq) )
mu.NotAfrica.mean <- apply( mu.NotAfrica , 2 , mean )
mu.NotAfrica.PI <- apply( mu.NotAfrica , 2 , PI , prob=0.97 )

#7.11
# plot African nations with regression
#overlay: posterior mean (MAP) regression line
#97% interval of regression line
d.A1 <- dd[dd$cont_africa==1,]
plot( log(rgdppc_2000) ~ rugged , data=d.A1 ,
col=rangi2 , ylab="log GDP year 2000" ,
xlab="Terrain Ruggedness Index" )
mtext( "African nations" , 3 )
lines( rugged.seq , mu.Africa.mean , col=rangi2 )
shade( mu.Africa.PI , rugged.seq , col=col.alpha(rangi2,0.3) )
# plot non-African nations with regression
d.A0 <- dd[dd$cont_africa==0,]
plot( log(rgdppc_2000) ~ rugged , data=d.A0 ,
col="black" , ylab="log GDP year 2000" ,
xlab="Terrain Ruggedness Index" )
mtext( "Non-African nations" , 3 )
lines( rugged.seq , mu.NotAfrica.mean )
shade( mu.NotAfrica.PI , rugged.seq )

#7.12
precis(m7.5)
#have to get a point-estimate gamma by hand (bR + bAR(1) in Africa, bR + bAR(0) outside)

#7.13
#process samples from the posterior to compute the posterior distribution of gamma
post <- extract.samples( m7.5 )
gamma.Africa <- post$bR + post$bAR*1
gamma.notAfrica <- post$bR + post$bAR*0

#7.14
#get means of these distributions: same as 7.12
# ~= MAP values
mean( gamma.Africa)
mean( gamma.notAfrica )

#7.15
#plot full distributions of slopes by region -- on one axis to visualize overlap
#inside Africa: blue
dens( gamma.Africa , xlim=c(-0.5,0.6) , ylim=c(0,5.5) ,
xlab="gamma" , col=rangi2 )
dens( gamma.notAfrica , add=TRUE )

#7.16 
#compute probability that slope within Africa is less than slope outside?
#compute difference between slopes for each sample in posterior
diff <- gamma.Africa - gamma.notAfrica
#what proportion of differences below zero?
sum( diff < 0 ) / length( diff )
#so, highly unlikely that slope associating ruggedness with log-GDP is lower inside Africa

#7.17 
#try reverse interpretation
# get minimum and maximum rugged values
q.rugged <- range(dd$rugged)
# compute lines and confidence intervals
mu.ruggedlo <- link( m7.5 ,
data=data.frame(rugged=q.rugged[1],cont_africa=0:1) )
mu.ruggedlo.mean <- apply( mu.ruggedlo , 2 , mean )
mu.ruggedlo.PI <- apply( mu.ruggedlo , 2 , PI )
mu.ruggedhi <- link( m7.5 ,
data=data.frame(rugged=q.rugged[2],cont_africa=0:1) )
mu.ruggedhi.mean <- apply( mu.ruggedhi , 2 , mean )
mu.ruggedhi.PI <- apply( mu.ruggedhi , 2 , PI )
# plot it all, splitting points at median
med.r <- median(dd$rugged)
ox <- ifelse( dd$rugged > med.r , 0.05 , -0.05 )
plot( dd$cont_africa + ox , log(dd$rgdppc_2000) ,
col=ifelse(dd$rugged>med.r,rangi2,"black") ,
xlim=c(-0.25,1.25) , xaxt="n" , ylab="log GDP year 2000" ,
xlab="Continent" )
axis( 1 , at=c(0,1) , labels=c("other","Africa") )
lines( 0:1 , mu.ruggedlo.mean , lty=2 )
shade( mu.ruggedlo.PI , 0:1 )
lines( 0:1 , mu.ruggedhi.mean , col=rangi2 )
shade( mu.ruggedhi.PI , 0:1 , col=col.alpha(rangi2,0.25) )

```

##7.18
Examine continuous interactions: slope varies continuously with continuous variable
```{r}
library(rethinking)
#blooms of tulips grown in different conditions
data(tulips)
d <- tulips
#water and shade predictor vars
#blooms - outcome variable to predict
str(d)
#7.19
m7.6 <- map(
alist(
blooms ~ dnorm( mu , sigma ) ,
mu <- a + bW*water + bS*shade ,
a ~ dnorm( 0 , 100 ) ,
bW ~ dnorm( 0 , 100 ) ,
bS ~ dnorm( 0 , 100 ) ,
sigma ~ dunif( 0 , 100 )
) ,
data=d )
m7.7 <- map(
alist(
blooms ~ dnorm( mu , sigma ) ,
mu <- a + bW*water + bS*shade + bWS*water*shade ,
a ~ dnorm( 0 , 100 ) ,
bW ~ dnorm( 0 , 100 ) ,
bS ~ dnorm( 0 , 100 ) ,
bWS ~ dnorm( 0 , 100 ) ,
sigma ~ dunif( 0 , 100 )
) ,
data=d )

#7.20
#fix error messages
m7.6 <- map(
alist(
blooms ~ dnorm( mu , sigma ) ,
mu <- a + bW*water + bS*shade ,
a ~ dnorm( 0 , 100 ) ,
bW ~ dnorm( 0 , 100 ) ,
bS ~ dnorm( 0 , 100 ) ,
sigma ~ dunif( 0 , 100 )
) ,
data=d ,
#use a different method to climb the posterior distribution
method="Nelder-Mead" ,
#increase max number of iterations (so optim searches longer)
control=list(maxit=1e4) )
m7.7 <- map(
alist(
blooms ~ dnorm( mu , sigma ) ,
mu <- a + bW*water + bS*shade + bWS*water*shade ,
a ~ dnorm( 0 , 100 ) ,
bW ~ dnorm( 0 , 100 ) ,
bS ~ dnorm( 0 , 100 ) ,
bWS ~ dnorm( 0 , 100 ) ,
sigma ~ dunif( 0 , 100 )
) ,
data=d ,
method="Nelder-Mead" ,
control=list(maxit=1e4) )

#7.21 
#look at the estimates
coeftab(m7.6,m7.7)

#7.22
#evidence that the interaction model is better:
compare( m7.6 , m7.7 )

#7.23
#centered prediction variable: mean is zero
#can understand model better with coefficients alone
d$shade.c <- d$shade - mean(d$shade)
d$water.c <- d$water - mean(d$water)

#7.24
#now: fix models by using centered variables
#and explicit start lists for better priors
m7.8 <- map(
alist(
blooms ~ dnorm( mu , sigma ) ,
mu <- a + bW*water.c + bS*shade.c ,
a ~ dnorm( 130 , 100 ) ,
bW ~ dnorm( 0 , 100 ) ,
bS ~ dnorm( 0 , 100 ) ,
sigma ~ dunif( 0 , 100 )
) ,
data=d ,
start=list(a=mean(d$blooms),bW=0,bS=0,sigma=sd(d$blooms)) )
m7.9 <- map(
alist(
blooms ~ dnorm( mu , sigma ) ,
mu <- a + bW*water.c + bS*shade.c + bWS*water.c*shade.c ,
a ~ dnorm( 130 , 100 ) ,
bW ~ dnorm( 0 , 100 ) ,
bS ~ dnorm( 0 , 100 ) ,
bWS ~ dnorm( 0 , 100 ) ,
sigma ~ dunif( 0 , 100 )
) ,
data=d ,
start=list(a=mean(d$blooms),bW=0,bS=0,bWS=0,sigma=sd(d$blooms)) )
coeftab(m7.8,m7.9)

#7.25
#compute predicted effect of increasing water by 1 unit
k <- coef(m7.7)
k[1] + k[2]*2 + k[3]*2 + k[4]*2*2

#7.26
#same if you use the centered interaction model
k <- coef(m7.9)
k[1] + k[2]*0 + k[3]*0 + k[4]*0*0

#7.27
#table of improved, centered estimates
precis(m7.9)

#7.28
#tryptych plot: 3 plots to show 2-way interaction between 2 continuous variables
# make a plot window with three panels in a single row 
par(mfrow=c(1,3)) # 1 row, 3 columns
# loop over values of water.c and plot predictions
shade.seq <- -1:1
for ( w in -1:1 ) {
dt <- d[d$water.c==w,]
plot( blooms ~ shade.c , data=dt , col=rangi2 ,
main=paste("water.c =",w) , xaxp=c(-1,1,2) , ylim=c(0,362) ,
xlab="shade (centered)" )
mu <- link( m7.9 , data=data.frame(water.c=w,shade.c=shade.seq) )
mu.mean <- apply( mu , 2 , mean )
mu.PI <- apply( mu , 2 , PI , prob=0.97 )
lines( shade.seq , mu.mean )
lines( shade.seq , mu.PI[1,] , lty=2 )
lines( shade.seq , mu.PI[2,] , lty=2 )
}

#7.29
#model with interaction: as design formula
m7.x <- lm( y ~ x + z + x*z , data=d )

#7.30
#same model, abbreviated
m7.x <- lm( y ~ x*z , data=d )

#7.31
#can explicitly subtract main effects from the model
m7.x <- lm( y ~ x + x*z - z , data=d )

#7.32
#design formula including main effects, 2-way interactions and the 3-way intx
m7.x <- lm( y ~ x*z*w , data=d )

#7.33
#expand lm's design formula
#The function model.matrix takes the design formula, absent the outcome variable, and expands it into a full set of variables for a linear model.
x <- z <- w <- 1
colnames( model.matrix(~x*z*w) )
```


