---
output: pdf_document
---
# Statistical Rethinking Chapter 3 problems

__Name: Nicole E Soltis__

## 3E1
```{r}
p_grid <- seq( from=0 , to=1 , length.out=1000 )
prior <- rep( 1 , 1000 )
likelihood <- dbinom( 6 , size=9 , prob=p_grid )
posterior <- likelihood * prior
posterior <- posterior / sum(posterior)
set.seed(100)
samples <- sample( p_grid , prob=posterior , size=1e4 , replace=TRUE )
```
library(rethinking
)
How much posterior probability lies below p = 0.2?
```{r}
plot(samples)
library(rethinking)
dens(samples)
sum( posterior[ p_grid < 0.2 ] )
```

## 3E2
How much posterior probability lies above p = 0:8?
```{r}
sum( posterior[ p_grid > 0.8 ] )
```

## 3E3
How much posterior probability lies between p = 0.2 and p = 0.8?
```{r}
sum( samples > 0.2 & samples < 0.8 ) / 1e4
```

## 3E4
20% of the posterior probability lies below which value of p?
```{r}
quantile( samples , 0.2 )
```

## 3E5
20% of the posterior probability lies above which value of p?
```{r}
quantile( samples , 0.8 )
```

## 3E6
Which values of p contain the narrowest interval equal to 66% of the posterior probability?
```{r}
library(rethinking)
HPDI( samples , prob=0.66 )
```

## 3E7
Which values of p contain 66% of the posterior probability, assuming equal posterior probability
both below and above the interval?

```{r}
library(rethinking)
PI( samples , prob=0.66 )
```

## 3M1
Suppose the globe tossing data had turned out to be 8 water in 15 tosses. Construct the posterior
distribution, using grid approximation. Use the same flat prior as before.

```{r}
p_grid <- seq( from=0 , to=1 , length.out=1000 )
prior <- rep( 1 , 1000 )
likelihood <- dbinom( 8 , size=15 , prob=p_grid )
posterior <- likelihood * prior
posterior <- posterior / sum(posterior)
```

## 3M2
Draw 10,000 samples from the grid approximation from above. Then use the samples to calculate
the 90% HPDI for p.

```{r}
set.seed(100)
samples <- sample( p_grid , prob=posterior , size=1e4 , replace=TRUE )
library(rethinking)
dens(samples)
HPDI( samples , prob=0.9 )
```

_STOP AFTER 3M2 FOR 02/25 ASSIGNMENT_

## 3M3

## 3M4

## 3M5

## 3H1

## 3H2

## 3H3

## 3H4

## 3H5